Comparison of long and short amplicon metabarcoding for determination of fungal species diversity and spatial turnover

### 2019-08-13 writing/MSA2019_presentation.Rnw
add pictures

% # Metabarcoding
% 
%   - Cloning + Sanger -> 454 -> Illumina
%   - For fungi, usually ITS2
%       - short
%       - intraspecific variation in most groups
%       - good database support (Unite 8.0 - 817 130 fungal sequences)
%       - ITS is "official barcode for fungi" \cite{schoch2012}
%       - fungal specific primers
%   - problems
%       - not alignable
%       - cannot place unknown sequences
%       - variant 5.8S
%       - variable length
%   
%   - LSU (26S/28S)
%       - alternating highly conserved domains and variable regions
%       - widely used for phylogenies
%       - universal primers
%       - less variable length
%       - more conserved than ITS
%       - not as good database support (RDP 11 fungal LSU - 125 525, Silva 132 LSU 907 382 of which )
%   - SSU (18S)
%       - alternating highly conserved and variable regions
%       - more conserved than LSU
%       - used for barcoding prokaryotes and "protists"
%       - good database support (Silva SSU), but not as good for fungi

@@

\begin{frame}{Upper Ouémé Forest Reserve}
\framesubtitle{Permanent study plots}
\vskip1.5em
\begincols
  \begincol{0.55\linewidth}
    \begin{figure}
      \begin{overprint}
      \onslide<1>
        \includegraphics{soudanian_benin}
      \onslide<2>
        \includegraphics{soudanian_benin_oueme}
      \onslide<3->
        \includegraphics{soudanian_benin_oueme_village}
      \end{overprint}
      \caption{Location of study sites in Bènin.}
    \end{figure}
  \endcol

  \begincol{0.45\linewidth}
  \begin{itemize}
    \tightlist
    \item<1-|alert@1> West Soudanian woodlands in central Bènin
    \item<1-|alert@1> 10° N Latitude
    \item<3-|alert@3> Two sites located near the villages Angaradebou and Gando
    \item<4-|alert@4> Dominated by \emph{Isoberlinia doka} (Caesalpinoideae; ECM host)
  \end{itemize}
  \endcol
\endcols
\end{frame}

@@

\begin{frame}{Sequencing}
  \framesubtitle{Which region?}
  \small
  \begincols
  \begincol{0.5\linewidth}
    \begin{block}{ITS2 (gITS7-ITS4)}
        \begin{itemize}
          \tightlist
          \item Internally transcribed spacer between 5.8S and 26/28S (LSU) subunits of ribosomal RNA
          \item Usually 250--350 bp.
          \item ITS --- standard DNA barcode for fungi.
          \item Relatively good species-level resolution, and good database support.
          \item Cannot be aligned across larger taxonomic groups.
          \item Groups with variant (or absent) 5.8S not sequenced.
        \end{itemize}
      \end{block}
    \endcol
    \begincol{0.5\linewidth}
      \begin{block}{Full ITS + partial LSU (ITS1-LR5)}
        \begin{itemize}
          \tightlist
          \item Full ITS plus the 5' end of LSU; 1500-2000 bp.
          \item ITS -- resolution to separate closely related taxa.
          \item LSU -- can be aligned across fungi.
          \item Primers are more universal than gITS7.
          \item Long amplicon with alternating variable and conserved regions --- susceptible to chimera formation and breakage.
        \end{itemize}
      \end{block}
    \endcol
  \endcols
\end{frame}

>>>>

\setbeamersize{text margin left=5mm,text margin right=5mm}

\begin{frame}{rDNA Regions and Databases}
\framesubtitle{rDNA regions}
\vskip -1ex

\includegraphics[width=\linewidth]{rDNA}

\vfill

\begin{overprint}
\onslide<1>
\small
\begin{tabulary}{\linewidth}{r|L|L|L}
& SSU (16S/18S) & ITS & LSU (23S/26S/28S) \\ \midrule
Variation &
  Highly conserved w/variable regions &
  Variable (5.8S conserved) &
  Conserved w/variable regions \\ \noalign{\smallskip}
Uses & 
  Prokaryota, broad Eukaryota &
  Official barcode for Fungi &
  Phylogenetics \\ \noalign{\smallskip}
Alignable &
  Yes &
  No &
  Yes \\ \noalign{\smallskip}
Databases &
  Silva, RDP (Prokaryotes only) &
  Unite, Warcup &
  Silva, RDP (Fungi) \\ 
\end{tabulary}

\onslide<2>
\begin{columns}[t]
  \begincol{0.3\linewidth}
    \begin{block}{Silva 132 (2017)}
      \small
      \textbf{SSU Ref:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 2M seqs
        \item 700k "Ref NR 99" (clusters)
        \item 16k Fungi, 60k other Eukaryota, 625k Prokaryota
      \end{itemize}
      \textbf{LSU Ref:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 200k seqs
        \item 5k Fungi, 24k other Eukaryota, 170k Prokaryota
      \end{itemize}
    \end{block}
  \endcol
  \begincol{0.3\linewidth}
    \begin{block}{Unite 8.0 (2019)}
    \small
    \textbf{ITS:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 800k seqs
        \item 120k "species hypotheses" (clusters)
        \item 70k Fungi, 40k plants, 10k other Eukaryota
        \itemsome sequences include LSU
      \end{itemize}
    \end{block}
  \endcol
  \begincol{0.3\linewidth}
    \small
    \begin{block}{RDP 11 (2016)}
      \small
      \textbf{Fungal LSU training set:}\\
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 11k seqs
        \item 8k Fungi, 3k other Eukaryota
        \item annotated to genus
      \end{itemize}
      \textbf{SSU dataset:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item Prokaryotes only
      \end{itemize}
      \textbf{Warcup Fungal ITS training set:}
      \begin{itemize}
        \footnotesize
        \tightlist
        \item 18k seqs
        \item all Fungi
      \end{itemize}
    \end{block}
  \endcol
\end{columns}
\center The three databases don't use the same taxonomy!
\end{overprint}
\end{frame}

@@

\begin{frame}{Study site}
\vskip1.5em
\begincols
  \begincol{0.55\linewidth}
    \begin{figure}
      \begin{overprint}
      \onslide<1>
        \includegraphics{soudanian_benin}
      \onslide<2>
        \includegraphics{soudanian_benin_oueme}
      \end{overprint}
      \caption{Location of study sites in Bènin.}
    \end{figure}
  \endcol

  \begincol{0.45\linewidth}
  \begin{itemize}
    \tightlist
    \item West Soudanian woodlands in central Bènin
    \item 10° N Latitude
    \item Two sites 30km apart
    \item Dominated by \emph{Isoberlinia doka} (Caesalpinoideae; ECM host)
    \item Many AM trees and shrubs, plus three more ECM.
    \item Fire \& termites
  \end{itemize}
  \endcol
\endcols
\end{frame}

\begin{frame}{Soudanian Woodlands}
  \begin{overprint}
    \onslide<1>
      \includegraphics{Foret_Oueme_wide}
    \onslide<2>
      \includegraphics{woodland}
  \end{overprint}
\end{frame}

@@

\begin{frame}{Sequencing}
  \framesubtitle{Which region?}
  \small
  \includegraphics[width=\linewidth]{rDNA_amplicons}

  \begincols
  \begincol{0.5\linewidth}
    \begin{block}{ITS2 (gITS7-ITS4)}
        \begin{itemize}
          \tightlist
          \item Internally transcribed spacer between 5.8S and 26/28S (LSU) subunits of ribosomal RNA
          \item Usually 250--350 bp.
          \item ITS --- standard DNA barcode for fungi.
          \item Relatively good species-level resolution, and good database support.
          \item Cannot be aligned across larger taxonomic groups.
          \item Groups with variant (or absent) 5.8S not sequenced.
        \end{itemize}
      \end{block}
    \endcol
    \begincol{0.5\linewidth}
      \begin{block}{Full ITS + partial LSU (ITS1-LR5)}
        \begin{itemize}
          \tightlist
          \item Full ITS plus the 5' end of LSU; 1000-2000 bp.
          \item ITS -- resolution to separate closely related taxa.
          \item LSU -- can be aligned across Fungi.
          \item Primers are more universal than gITS7.
          \item Long amplicon with alternating variable and conserved regions --- susceptible to chimera formation and breakage.
        \end{itemize}
      \end{block}
    \endcol
  \endcols
\end{frame}
